Analyzing the effects of COVID-19 related social media use on well-being

### 2021-12-13 manuscript.Rmd
- updated paper for submission

---
title             : "No meaningful effects of COVID-19 related social media use on well-being"
shorttitle        : "COVID-19 related social media use and well-being"

author: 
  - name          : "Blinded for review"
    affiliation   : "1"
    corresponding : yes
    address       : "Blinded for review"
    email         : "Blinded for review"

affiliation:
  - id            : "1"
    institution   : "Blinded for review"

authornote: |
  Blinded for review.

abstract          : "In times of crisis such as the Corona pandemic, it is important that citizens stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-19 related topics might put personal well-being at risk. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects cross lagged panel models, controlling for several stable and varying covariates. Results showed that the effects of various types of COVID-19 related social media use on several indicators of well-being were very small, arguably too small to matter. The findings suggest fears that social media use during times of crisis might be detrimental for well-being are likely to be unfounded."
keywords          : "COVID-19, Coronavirus, well-being, affect, life satisfaction, social media use, news use, communication, random effects within between model, panel study, longitudinal."

@@

documentclass     : "apa6"
classoption       : "man"
# output            : papaja::apa6_pdf
# output            : papaja::apa6_word
header-includes:
  - \setlength{\parskip}{0em}
  - \raggedbottom
  - \note{\clearpage}
---

```{r analysis-preferences, include=F, cache=F}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      cache=T,
                      echo=F
                      # , fig.pos = 'h'
                      )
```

@@

# make project reproducible
# devtools::install_github("rstudio/renv")  # install renv
# renv::init()  # initialize renv

# install necessary packages
packages <- c("devtools", "english", "ggplot2", "kableExtra", 
              "lavaan", "lme4", "lmerTest",
              "magrittr", "semPlot", "tidyverse")
# renv::install(packages)
# renv::install("tdienlin/td@v.0.0.2.5")  # uncomment to install
# renv::install("crsh/papaja@v0.1.0.9997")  # uncomment to install
# renv::install("crsh/papaja@devel")  # uncomment to install

# load packages
lapply(c(packages, "papaja", "td"), library, character.only = TRUE)

# save snapshot of installed packages
# renv::snapshot()

# load workspace created from analyses.rmd to use results
load("data/workspace.RData")

# export references
# r_refs(file = "bibliography/r-references.bib")
```

During the COVID-19 pandemic, 
<!-- and especially in the beginning,  -->
numerous events unfolded in quick succession.
Several open questions emerged.
How dangerous is the virus? 
Is it spreading in my region? 
How is it transmitted? 
How can I protect myself?
Because for many it was (and continues to be) a matter of life or death, citizens had to stay informed regarding the latest developments.
Governments around the world implemented safety measures, ranging from wearing masks, keeping physical distance, to complete lockdowns.
In this extraordinary situation, people used media excessively to attain information, and especially social media were at an all time high [@statistaAverageDailyTime2021].
A new phenomenon termed "doomscrolling" emerged:
People could not stop using social media to learn about COVID-19 related news.

Several people reported that they were glued to their screens and could not pursue other relevant activities such as working, taking a break, or even child-care [@kleinDarklySoothingCompulsion2021].
As doomscrolling increased it became doubtful whether such a surge in social media use could still be considered useful and adaptive, or whether it created an additional and new psychological danger for users' mental health [@sandstromDoomscrollingCOVIDNews2021].
These concerns seem justified: A study with 6,233 people from Germany found that "[f]requency, duration and diversity of media exposure were positively associated with more symptoms of depression and unspecific and COVID-19 specific anxiety" [@bendauAssociationsCOVID19Related2021], which implies negative consequences for well-being.

As a result, with this study I want to build on this research and investigate the question whether COVID-19 related social media use during the pandemic affected well-being.
To this end, I analyze a large-scale panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of 24 waves and has an overall sample size of `r nrow(d_wide)`.
At least 1,500 participants took part per wave, and the sample is representative of the Austrian population.
The panel study collected a large number of variables.
Being able to control for many confounding third variables, both stable and varying, and together with the longitudinal design, this creates a unique opportunity to investigate the causal effects of COVID-19 related social media use on well-being.

## Defining Well-being and Media Use

The underlying theories that guided the selection of variables for this study are the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated medation [@meierComputerMediatedCommunicationSocial2020].
According to the two-continua model, mental health consists of two dimensions: psychopathology and well-being.
<!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
Well-being can be further differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018].
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being focuses on eudaimonic aspects like fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, life satisfaction should be considered a meta concept that combines psychological and subjective well-being, because it represents a general appraisal of one's life.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020].
<!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
To capture well-being in this study, I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); sixth, the message (e.g., content) [@meierComputerMediatedCommunicationSocial2020].
Whereas the first four levels are focused on the _channel_, the last two address the _type_ of communication.
To measure social media use for consumption of COVID-19 related news and topics, I here employ both the channel and the communication perspective, which together should provide a more nuanced understanding of communication.

First, I will investigate how different types of communication affect well-being.
Specifically, I will differentiate between active and passive use.
I will distinguish (a) _reading_ COVID-19 related social media news (passive), _posting_ content regarding COVID (active), and _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I will analyze also how using the most prominent branded applications affect well-being, and whether this effect changes across applications.
According to @meierComputerMediatedCommunicationSocial2020, branded apps are separate entities with different effects. 
Therefore, Twitter might have a different effect than WhatsApp because of their different affordances.
For example, @waterlooNormsOnlineExpressions2018 found that it is more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. 
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about _general_ social media during times of COVID, but on social media use _focused_ on COVID-19 related news and interactions. 
Examples of such media use include posting about the pandemic or retweeting COVID-19 related posts.

@@

So far, there is only little empirical research on COVID-19 related social media use and well-being.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
Hence, this might hint at potential negative effects of social media as news use.
However, note that this finding comes from a between-person relation stemming from cross-sectional data (see below).
We, therefore, don't know whether the differences in mental health and well-being are due to social media use or other third variables, such as age, health, or education.
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, and found both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.

On the other hand, the question of whether and how _general_ social media use affect well-being is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses), found that the relation between social media use and well-being is likely in the negative spectrum but very small, potentially too small to matter [@meierComputerMediatedCommunicationSocial2020].
What determines whether an effect is considered trivial or small?
If we refer to standardized effect sizes, according to @cohenPowerPrimer1992 small effect sizes start at _r_ = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputerMediatedCommunicationSocial2020; @fergusonThisMetaanalysisScreen2021].

These overviews are well aligned with several individual new studies that employed advanced methods [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020].
For example, @beyensSocialMediaUse2021 reported that although for some users (roughly one quarter) the effects were negative, for almost the same number of users they were positive, while for the majority the effects were neutral.
In conclusion, most effects are somewhere between trivial and small.
I therefore think it's most convincing to expect trivial to small effects also in the case of COVID-19 related social media use.

From a theoretical perspective, how could we explain how COVID-19 related social media use might affect well-being?
In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect. 
In advance, there does not seem to be a clear winner, and it's likely that both positive and negative effects are equally strong.

First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which is more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines across many countries worldwide.
If one learns about such things, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news might be depressing, perhaps even changing some general perspectives on life, without any further mediating processes.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and potentially uplifting news, potentially compensating for the negative news.
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

There could also be _indirect_ effects.
When doomscrolling, users are captivated to such an extent that they cannot stop using social media.
For example, during the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021].
As has been expressed by many before, it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020].
Overuse, however, is more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017].
To explain, overuse likely impairs well-being if it replaces other meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media, there is potential for negative effects.
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

On the other hand, one can make the case that overuse can also be beneficial, especially in times of a pandemic, even if the use is mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at that time.
In situations where meaningful and functional activities are explicitly or implicitly prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
In fact, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand information on the current developments.
<!-- On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being. -->

Together, the strongest argument seems to be that _in general_ the effects of social media on well-being are, on average, small at best.
Because this study looks at only _one part_ of social media use---namely, COVID-19 related interactions---it is even more focused, and the overall potential of the effects should diminish even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->
Therefore, I expect to find that COVID-19 related communication on social media will not affect well-being in a meaningful or relevant way.

@@

Testing this hypothesis, however, is not trivial.
First, in contrast to most hypotheses typically posited in the social sciences, it implicitly contains an effect size, a so-called smallest effect size of interest (SESOI).
Effectively testing this hypothesis necessitates defining what's considered a "trivial effect size" and what's not.
Above I already referred to standardized effect sizes.
However, standardized effect sizes should only be a first step toward evaluating an effect's relevance [@baguleyStandardizedSimpleEffect2009].
Standardized effect sizes are determined by a sample's variance.[^cohensd]
However, this is problematic: The question of whether or not social media use affects a particular person in a relevant way should not depend on the variance in the sample in which that person's data were collected.
Instead, it should depend on absolute criteria.
What could be a minimally interesting, a nontrivial effect?
Because this is a normative and ultimately philosophical question, there can never be a clear, single, or unanimous answer. 
In the end, it is a personal question.
However, it is still both necessary and fruitful to try to answer it. 
I suggest the following SESOI for this research question:

@@

If a heavy user of COVID-19 related social media news suddenly _completely stops_ using social media, this should have a _noticeable_ impact on their overall well-being.

What would this mean practically and how could it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, the example from above would imply that a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably differentiate between _seven_ levels of satisfaction with health.
So we could state that a four unit change in social media use should result in a one unit change in life satisfaction, if satisfaction is measured on a 7-point scale.
Statistically, in a regression _b_ estimates the change in the dependent variable if the independent variable increases by one point.
Transferred to our example, we would hence expect a SESOI of _b_ = 1 / 4 = .25.

In this study, life satisfaction was measured on an 11-point scale, which is more than the seven degrees people can reliably differentiate. 
We hence need to transform a 1-point change on a 7-point scale to its equivalent on an 11-point scale, which is 11 / 7 = `r 11/7`.
Hence, we now expect our effect to be at least as large as _b_ = .25 * `r 11/7` = `r 11/7*.25`.
For affect, which was measured on a 5-point scale, our SESOI would therefore be _b_ = 1/4 * 5/7 = `r 5 / 7 * .25`.
Because we are agnostic as to whether the effects are positively or negatively nontrivial, the null region will include negative and positive effects (in Bayesian terms, this is called the region of practical equivalence [ROPE]).
In order not to exaggerate precision, and to be a bit less conservative, these numbers will be reduced to nearby thresholds. 
Note that other researchers also decreased or recommended decreasing thresholds for effect sizes when analyzing within-person or cumulative effects (@beyensSocialMediaUse2021; @funderEvaluatingEffectSize2019).
Together, this leads to an indifference region ranging from _b_ = -.30 to _b_ = .30 for life satisfaction, and _b_ = -.15 to _b_ = .15 for positive and negative affect.

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs help investigate causality.
Using longitudinal designs alone, however, is not sufficient for correct causal statements.
In addition, we also need to control for third variables, and importantly also for _varying_ third variables.

Consider the following example.
Imagine that a person suddenly starts using social media much more than usual, and then at the same time also becomes less satisfied with their life.
After some time, use recedes again, whereas life satisfaction returns to prior levels.
If this happens to several people at the same time, in a longitudinal study we would then observe a significant effect of social media use on life satisfaction.
However, it could also be the case that during the study there was a major exogenous event (say, a pandemic), which is why a large part of the working population lost their jobs.
Hence, the causal effect reported above was likely confounded, because in reality it was the pandemic that caused both social media use to rise and life satisfaction to plummet.

Thus, only if we can control for _all_ potential varying third variables, can we correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure that we have included all varying covariates, which makes absolute statements regarding causality impossible.
In addition, when determining the overall causal effect, we should _not_ control for mediating variables [@rohrerThinkingClearlyCorrelations2018].
Doing so would bias and distort our assessment of the role of social media use.
Complicating matters, often it is unclear if a variable is a mediator or a confounder.[^collider]
However, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many _relevant_ varying and nonvarying third variables as possible, while knowing that absolute certainty regarding causality cannot be reached.

@@

When looking for suitable control candidates, ideally we select variables that affect both media use and well-being. 
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect _only_ well-being. 
However, in doing so not much is gained or lost because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016].[^mcelreath]

[^mcelreath]: But see @mcelreathYesterdayClass2021.

In this study, I hence plan to control for the following variables, which either have already been shown to affect both social media use and well-being, or which are very likely to do so, and which are likely not mediators:
<!-- (I'll additionally include variables that likely affect only well-being, also to obtain a comparison benchmark for social media effects): -->
gender, age, education, Austria country of birth, Austria country of birth of parents, text-based news consumption, video-based news consumption, residency Vienna, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, satisfaction with democracy, disposition to take risks, and locus of control.
I will not control for variables such as trust in institutions or media, because these variables might be influenced by social media use to a meaningful extent.

Next to including covariates, it is now increasingly understood that causal effects need to be analyzed from an internal, within-person perspective.
If a specific person changes their media diet, we need to measure how this affects _their_ well-being.
Between-person comparison from cross-sectional data, where participants are interviewed only once, cannot provide such insights.
In this study, I will hence differentiate between within-person effects and between-person relations. 
As explicated in the hypothesis, to test the hypothesis I will thus consider only the within-person effects.

>>>>

---
title             : "Analyzing the effects of COVID-19 related social media use on well-being"
shorttitle        : "COVID-19 related social media use and well-being"

author: 
  - name          : "Name blinded for review"
    affiliation   : "1"
    corresponding : yes
    address       : ""
    email         : ""

affiliation:
  - id            : "1"
    institution   : ""

abstract          : "In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-19 related reasons affects psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded."
keywords          : "COVID-19, Coronavirus, well-being, affect, life satisfaction, social media use, news use, communication, random effects within between model, panel study, longitudinal study."

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes:
  - \setlength{\parskip}{0em}
  - \raggedbottom
  - \note{\clearpage}
---

```{r analysis-preferences, include=F, cache=F}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      cache=T, echo=F)
```

@@

# install necessary packages
# devtools::install_github("https://github.com/tdienlin/td@v.0.0.2.5")
# devtools::install_github("https://github.com/crsh/papaja@devel")

# load packages
packages <- c("devtools", "english", "ggplot2", "kableExtra", 
              "lavaan", "lme4", "lmerTest",
              "magrittr", "semPlot", "tidyverse")
lapply(c(packages, "papaja", "td"), library, character.only = TRUE)

# load workspace created from analyses.rmd to use results
load("data/workspace_2.RData")
```

During the COVID-19 pandemic, 
numerous events unfolded in quick succession and several open questions emerged.
How dangerous is the virus? 
Is it spreading in my region? 
How is it transmitted, and how can I protect myself?
Because for many it was (and still is) a matter of life or death, people aimed to stay informed regarding the latest developments.
Governments around the world implemented safety measures, such as wearing masks, keeping physical distance, or enforcing lockdowns.
In this extraordinary situation, many people used media excessively to attain information, and especially social media were at an all time high [@statistaAverageDailyTime2021].

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed "doomscrolling":
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children [@kleinDarklySoothingCompulsion2021].
As doomscrolling increased it became doubtful whether such social media is helpful, or whether it created an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
These concerns seem justified: 
A study with 6,233 people from Germany conducted during the pandemic found that "[f]requency, duration and diversity of media exposure were positively associated with more symptoms of depression" [@bendauAssociationsCOVID19Related2021, p. 283].

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of 24 waves and has an overall sample size of `r nrow(d_wide) %>% prettyNum(big.mark = ",")` participants.
<!-- At least 1,500 participants took part per wave, and the sample is representative of the Austrian population. -->
The panel study collected a large number of psychological and demographic variables.
<!-- Being able to control for many confounding third variables, both stable and varying, together with the longitudinal design, this creates a unique opportunity  -->
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

## Understanding Well-being and Media Use

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a].
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
<!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018].
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020].
<!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together provides a nuanced understanding of communication.

First, I investigate how different types of communication affect well-being.
Specifically, I differentiate between active and passive use.
I distinguish (a) _reading_ (passive), (b) _posting_ (active), and (c) _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects. 
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. 
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content. 
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

@@

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations [@sewallObjectivelyMeasuredDigital2021].
The results showed that digital technology did not have any significant effects on mental health [for a similar study with comparable results, see @bradleyStressMoodSmartphone2021].
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use [see also @dornemannHowGoodBad2021; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

The question of whether and how social media use affects well-being _in general_, on the other hand, is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small [@meierComputermediatedCommunicationSocial2020a]---potentially too small to matter.
What determines whether or not an effect is considered small or trivial?
As a starting point, we could refer to standardized effect sizes.
According to @cohenPowerPrimer1992, small effect sizes start at _r_ = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021].

These overviews are well aligned with several individual studies employing advanced methods [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020].
For example, @beyensSocialMediaUse2021 reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
In conclusion, most effects are likely somewhere between trivial and small.
I therefore expect that also in the case of COVID-19 related social media use effects will be trivial to small.

From a theoretical perspective, how could we explain whether COVID-19 related social media use might affect well-being?
In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent.
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.

First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021].
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

There could also be _indirect_ effects.
When doomscrolling, users are captivated to such an extent that they cannot stop using social media.
For example, during the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021].
In general, as has been expressed by many before, it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020].
Overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017].
To explain, overuse likely impairs well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media when doomscrolling, there is potential for negative effects.
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

On the other hand, one can make the case that overuse might also be beneficial, especially in times of a pandemic---even if the use is mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.
<!-- On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being. -->

Together, the strongest argument to me is that _in general_ the effects of social media on well-being are, on average, small at best.
Because this study only looks at _one part_ of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->
Therefore, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

Testing this hypothesis, however, is not trivial.
First, in contrast to most hypotheses typically posited in the social sciences it implicitly contains an effect size, a so-called smallest effect size of interest (SESOI).
Effectively testing this hypothesis necessitates defining what's considered a "trivial effect size" and what's not.
Above I already referred to standardized effect sizes.
However, standardized effect sizes should only be a first step toward evaluating an effect's relevance [@baguleyStandardizedSimpleEffect2009].
Standardized effect sizes are determined by a sample's variance,[^cohensd] which is problematic: 
The question of whether or not social media use affects a particular person in a relevant way should not depend on the variance in the sample in which that person's data were collected.
Instead, it should depend on absolute criteria.

What could be a minimally interesting, nontrivial effect?
Because this is a normative and ultimately philosophical question, there can never be a clear, single, or unanimous answer. 
<!-- In the end, it is a personal question. -->
However, it is still necessary and helpful to try provide such a plausible benchmark. 
I therefore suggest the following SESOI for this research question:

@@

> SESOI: If a heavy user of COVID-19 related social media news suddenly _stops_ using social media altogether, this should have a _noticeable_ impact on their overall well-being.

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably distinguish _seven_ levels of satisfaction with health.
So if satisfaction is measured on a 7-point scale, we would state that a four unit change in social media use should result in a one unit change in life satisfaction. 
(For more information, see Methods section "Inference Criteria.")

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements [@rohrerTheseAreNot2021].
In addition, we for example also need to control for confounding third variables, and importantly also for _varying_ third variables.

To illustrate, consider the following example.
Imagine that a person suddenly starts using social media much more than usual, and then after some time  becomes less satisfied with their life.
Eventually, use and life satisfaction return to prior levels.
If this happens to several people at the same time, in a longitudinal study we could then observe a significant effect of social media use on life satisfaction.
However, it could also be the case that during the study there was a major exogenous event (say, a pandemic), which caused large parts of the working population to loose their jobs.
Hence, the causal effect reported above was confounded, because in reality it was the pandemic that caused both social media use to rise and life satisfaction to go down.

Thus, only when controlling for _all_ relevant confounders, we can correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible.
In addition, when determining the overall causal effect, we need to make sure _not_ to control for mediating variables [@rohrerThinkingClearlyCorrelations2018], for doing so would bias our assessment of the causal effect.
Complicating matters further, it is often unclear if a variable is a mediator or a confounder.[^collider]
However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many varying and nonvarying confounders as possible (which I believe is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached.

@@

When searching for suitable candidates for confounders, we should look for variables that affect both media use and well-being. 
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect only social media use or well-being. 
However, in doing so not much is gained or lost, because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016; but see @mcelreathYesterdayClass2021].

In this study, I hence plan to control for the following variables, which either have already been shown to affect both social media use and well-being or which are likely to do so, and which also aren't mediators:
<!-- (I'll additionally include variables that likely affect only well-being, also to obtain a comparison benchmark for social media effects): -->
gender, age, education, Austria country of birth, Austria country of birth of parents, text-based news consumption, video-based news consumption, residency Vienna, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, satisfaction with democracy, disposition to take risks, and locus of control.
I will not control for variables such as trust in institutions or trust in media, because these variables might be influenced by social media use to a meaningful extent.

Next to including covariates, it's now increasingly understood that causal effects should be analyzed from an internal, within-person perspective [@hamakerWhyResearchersShould2014].
If a specific person changes their media diet, we need to measure how this behavior affects _their own_ well-being.
Between-person comparisons from cross-sectional data, where participants are interviewed only once, cannot provide such insights.
In this study, I hence differentiate between-person relations from within-person effects. 
And as explicated above, to test the hypothesis I thus consider only the within-person effects.
